---
title: "ESM 206 Lab 5"
author: "Janelle Christensen"
date: "10/29/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Lab 5 Objectives

- Learn to parse dates with 'lubridate'
- Get counts of observations with count()
- Uncount()
- One and two-sample t-test
- Created a heatmap with geom_tile()

```{r, include = FALSE}

# read in Data and attach packages

library(tidyverse)
library(here)
library(janitor)

lobster_abundance <- read_csv(here::here("data", "lobster_abundance.csv"),
                              na = "-99999") %>%
  janitor::clean_names()

# not in tidy format because each observation is not given its own row - rows have counts in them


```

Use tidyr::uncount() function to convert our lobster data from frequency format to case format

```{r}

lobster_tidy <- lobster_abundance %>% 
  tidyr::uncount(lobster_count) #this gets rid of the count column and then keeps all observations the same, expands it to the number of rows that each "lobster_count" counts

```

####Exploratory Data Visualizations

We're only going to consider "site" as our variable of interest

```{r}

ggplot(lobster_tidy, aes(x = site, y = size_mm)) +
  geom_jitter(aes(color = site),
              alpha = 0.5,
              width = 0.2)

# histogram

ggplot(lobster_tidy, aes(x = size_mm)) +
  geom_histogram(aes(fill = site)) +
  facet_wrap(~site, scales = "free") # be careful with scales = "free" because it will make it so that comparing across graphs is difficult to do and it can make us create irresponsible data

# q-q plot

ggplot(lobster_tidy, aes(sample = size_mm)) +
  geom_qq() +
  facet_wrap(~site)

```

do these look normally distributed? yes. Should we assume that lobsters are normally distributed? yes. This means that we can use the mean to evaluate these samples.

If your data are super skewed, you might want to compare the median as a central tendency rather than the mean. You want to know that the mean is a good indicator of central tendency before you start to compare it between populations

#### Convert the 'date' column to class 'Date'

We'll use the lubridate package to convert to date format and the easily parse month & year

```{r}

lobster_date <- lobster_tidy %>% 
  mutate(
    date_new = lubridate::mdy(date)
  ) #mdy means that this data is in the format of month date year, could be any variation of this dym, myd, etc

```

Now, let's parse year and month using lubridate::month() and lubridate::year()

```{r}

lobster_parse_date <- lobster_date %>% 
  mutate(
    obs_month = lubridate::month(date_new, label = TRUE), 
    obs_year = lubridate::year(date_new)
  )

```

Now let's find counts of observed lobsters based on different hierarchical groupings:

First, let's count by year and month

```{r}

lobster_ym <- lobster_parse_date %>% 
  dplyr::count(obs_year, obs_month)

lobster_ym

lobster_y <- lobster_parse_date %>% 
  dplyr::count(obs_year)

lobster_y

lobster_site <- lobster_parse_date %>% 
  dplyr::count(site)

lobster_site

```

If we want to create a summary table that contains statistics OTHER than counts by group, it's easier to use group_by() + n()

```{r}

lobster_summary <- lobster_parse_date %>% 
  group_by(site) %>% 
  summarize(
    mean_size = mean(size_mm, na.rm = TRUE),
    sd_size = sd(size_mm, na.rm = TRUE),
    lobster_number = n()
  )

lobster_summary

```

#### Find confidence intervals

Confidence interval tells us that if we took many samples, in 95% of the samples we took, we would expect to find the population mean (for a 95% confidence interval)

Use t.test() function to find confidence intervals (for one sample) and perform t-test to compare means of two samples (...this will be covered concepturally in lecture week 6)

```{r}

ivee_lobsters <- lobster_tidy %>% 
  filter(site == "IVEE") %>% 
  pull(size_mm)

t.test(ivee_lobsters)

```

#### Two-sample t-test to compare means

We want to ask: is there a significant difference in lobster lengths at Naples and Mohawk reefs?

We've done our neccesary exploratory analyses to determine that a 2-sample t-test for means comparison is appropriate.

```{r}

napl_sample <- lobster_tidy %>% 
  filter(site == "NAPL") %>% 
  pull(size_mm)

mohk_sample <- lobster_tidy %>% 
  filter(site =="MOHK") %>% 
  pull(size_mm)

mn_ttest <- t.test(mohk_sample, napl_sample)

mn_ttest

# the 95% confidence interval is negative in this because the sample I wrote first is smaller than the one I wrote second

```

There is a significant difference in lobster lengths between Naples and Mohawk Reef. (t(df) = statsistic, p < 0.001, alpha = 0.05).

```{r}

lobster_mn <- lobster_tidy %>% 
  filter(site %in% c("NAPL", "MOHK"))

mn_ttest2 <- t.test(size_mm ~ site, lobster_mn)

mn_ttest2$p.value
mn_ttest2$statistic

```

Here is the p-value: `r mn_ttest2$p.value` look at it! it shows up as a variable in my knitted document

Here is my t-statistic: `r mn_ttest2$statistic`

#### Now, a heatmap!

```{r}

lobster_ys <- lobster_parse_date %>% 
  count(obs_year, site)

ggplot(data = lobster_ys, aes(x = obs_year, y = site)) +
  geom_tile(aes(fill = n))

```

